---
title: "2. Analysis -- Exercise-Induced Transcriptomic Remodeling in an MPTP Mouse Model of Parkinson's Disease"
author: "Erika Li"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r libraries}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(EnhancedVolcano)
library(fgsea)
library(GSVA)
library(msigdbr)
library(ComplexHeatmap)
library(cowplot)
library(ashr)
library(ggrepel)
library(circlize)
library(org.Mm.eg.db)
library(DGEobj.utils)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(biomaRt)
library(broom)
library(clusterProfiler)
library(enrichplot)
library(AnnotationDbi)
library(stringr)
```

```{r load-data}
count_mat  <- readRDS("../clean_data/cleaned_counts_matrix.RDS")
counts_df  <- readRDS("../clean_data/cleaned_counts.RDS")
metadata   <- readRDS("../clean_data/cleaned_metadata.RDS")
```

---

# DESeq2

```{r deseq2-object}
dds <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = metadata,
  design    = ~ group
)
dds <- dds[rowSums(counts(dds)) > 10, ]
dds <- DESeq(dds)

dir.create("results", showWarnings = FALSE)
saveRDS(dds, "../results/dds_full_dataset.rds")
```

---

# Quality Control

## PCA

```{r pca, message=FALSE}
vsd <- vst(dds, blind = TRUE)
saveRDS(vsd, "../results/vsd_full_dataset.rds")

pca        <- plotPCA(vsd, intgroup = c("lesion", "exercise"), returnData = TRUE)
percentVar <- round(100 * attr(pca, "percentVar"))

p_pca <- ggplot(pca, aes(x = PC1, y = PC2, color = exercise, shape = lesion)) +
  geom_point(size = 6, alpha = 0.95) +
  scale_color_manual(
    values = c("no_exercise" = "#D55E00", "exercise" = "#0072B2"),
    labels = c("no_exercise" = "No Exercise", "exercise" = "Exercise")
  ) +
  scale_shape_manual(
    values = c("control" = 16, "MPTP" = 17),
    labels = c("control" = "Saline", "MPTP" = "MPTP")
  ) +
  labs(
    x     = paste0("PC1 (", percentVar[1], "%)"),
    y     = paste0("PC2 (", percentVar[2], "%)"),
    color = "Exercise",
    shape = "Lesion"
  ) +
  theme_classic(base_size = 18) +
  theme(
    legend.position  = "right",
    legend.title     = element_text(face = "bold"),
    axis.title       = element_text(face = "bold"),
    axis.text        = element_text(color = "black"),
    axis.line        = element_line(linewidth = 0.8),
    panel.border     = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

p_pca

dir.create("../results/figures", showWarnings = FALSE)
ggsave(
  filename = "../results/figures/poster_PCA_plot.png",
  plot     = p_pca,
  width    = 7, height = 5, units = "in", dpi = 300
)
```

## Sample-to-Sample Distance Heatmap

```{r distance-heatmap, message=FALSE}
sample_dists   <- dist(t(assay(vsd)))
sample_dist_mat <- as.matrix(sample_dists)
colnames(sample_dist_mat) <- gsub("_", "-", colnames(sample_dist_mat))
rownames(sample_dist_mat) <- gsub("_", "-", rownames(sample_dist_mat))

annotation_col <- metadata[, c("lesion", "exercise")]
rownames(annotation_col) <- metadata$sample
annotation_col$exercise <- factor(
  annotation_col$exercise,
  levels = c("no_exercise", "exercise"),
  labels = c("No exercise", "Exercise")
)
annotation_col$lesion <- factor(
  annotation_col$lesion,
  levels = c("control", "MPTP"),
  labels = c("Control", "MPTP")
)
colnames(annotation_col) <- c("Lesion", "Exercise")

annotation_colors <- list(
  Exercise = c("No exercise" = "#E89278", "Exercise" = "#56B4E9"),
  Lesion   = c("Control" = "grey70", "MPTP" = "grey20")
)

sample_order <- c(
  "MC1-1","MC1-2","MC1-3","MC1-4",
  "MC2-1","MC2-2","MC2-3",
  "MC3-1","MC3-2","MC3-3",
  "MC4-1","MC4-2"
)

dist_heatmap <- Heatmap(
  sample_dist_mat,
  name                     = "Distance",
  col                      = colorRamp2(c(0, 75, 150), c("#2C7BB6", "#FFFFBF", "#D7191C")),
  column_order             = sample_order,
  row_order                = sample_order,
  top_annotation           = HeatmapAnnotation(
    Exercise = annotation_col$Exercise,
    Lesion   = annotation_col$Lesion,
    col      = annotation_colors
  ),
  show_row_names           = TRUE,
  show_column_names        = TRUE,
  column_title             = "Sample-to-sample transcriptomic distances",
  clustering_method_rows   = "complete",
  clustering_method_columns = "complete"
)

png(
  filename = "../results/figures/sample_distance_heatmap_plot.png",
  width = 7 * 300, height = 5 * 300, res = 300
)
dist_heatmap
dev.off()

dist_heatmap
```

---

# Differential Expression

## Contrast Helper

```{r contrast-helper}
# define contrasts: numerator vs denominator
contrast_list <- list(
  MC2_vs_MC4 = c("group", "MPTP_no_exercise",    "control_no_exercise"),
  MC1_vs_MC2 = c("group", "MPTP_exercise",        "MPTP_no_exercise"),
  MC3_vs_MC4 = c("group", "control_exercise",     "control_no_exercise")
)

# returns raw (for fgsea) and ashr-shrunken (for plotting) results
run_contrast <- function(dds, contrast) {
  res_raw    <- results(dds, contrast = contrast)
  res_shrunk <- lfcShrink(dds, contrast = contrast, res = res_raw, type = "ashr")
  res_raw    <- res_raw[order(res_raw$pvalue), ]
  res_shrunk <- res_shrunk[order(res_shrunk$pvalue), ]
  list(raw = res_raw, shrunk = res_shrunk)
}
```

```{r run-contrasts}
MC1_MC2 <- run_contrast(dds, contrast_list$MC1_vs_MC2)
MC2_MC4 <- run_contrast(dds, contrast_list$MC2_vs_MC4)
MC3_MC4 <- run_contrast(dds, contrast_list$MC3_vs_MC4)
```

```{r gene-map}
# shared ensembl -> symbol map used throughout
gene_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys    = rownames(MC1_MC2$shrunk),
  keytype = "ENSEMBL",
  columns = "SYMBOL"
)

MC1_MC2$shrunk$gene_name <- gene_map$SYMBOL[
  match(rownames(MC1_MC2$shrunk), gene_map$ENSEMBL)]
MC1_MC2$shrunk$gene_name[is.na(MC1_MC2$shrunk$gene_name)] <-
  rownames(MC1_MC2$shrunk)[is.na(MC1_MC2$shrunk$gene_name)]

MC1_MC2$raw$gene_name <- gene_map$SYMBOL[
  match(rownames(MC1_MC2$raw), gene_map$ENSEMBL)]
MC3_MC4$raw$gene_name <- gene_map$SYMBOL[
  match(rownames(MC3_MC4$raw), gene_map$ENSEMBL)]
```

## MPTP vs Control (MC2 vs MC4)

```{r volcano-mc2-mc4}
# no genes pass adjusted p threshold at this chronic time point
EnhancedVolcano(
  MC2_MC4$shrunk,
  lab            = rownames(MC2_MC4$shrunk),
  x              = "log2FoldChange",
  y              = "padj",
  pCutoff        = 0.05,
  FCcutoff       = 1,
  title          = "MPTP vs Control (no exercise)",
  subtitle       = NULL
)
```

## MPTP + Exercise vs MPTP (MC1 vs MC2)

```{r volcano-mc1-mc2, fig.width=9, fig.height=6}
genes_to_label <- c("Fosl2", "Egr1", "Nr4a1", "Nr4a3")

keyvals <- ifelse(
  MC1_MC2$shrunk$padj < 0.05 & MC1_MC2$shrunk$log2FoldChange > 0, "red",
  ifelse(
    MC1_MC2$shrunk$padj < 0.05 & MC1_MC2$shrunk$log2FoldChange < 0, "blue",
    "grey85"
  )
)
names(keyvals) <- rep("NS", length(keyvals))
names(keyvals)[MC1_MC2$shrunk$padj < 0.05 & MC1_MC2$shrunk$log2FoldChange > 0] <- "Up"
names(keyvals)[MC1_MC2$shrunk$padj < 0.05 & MC1_MC2$shrunk$log2FoldChange < 0] <- "Down"

volcano_MC1_MC2 <- EnhancedVolcano(
  MC1_MC2$shrunk,
  lab              = MC1_MC2$shrunk$gene_name,
  selectLab        = genes_to_label,
  x                = "log2FoldChange",
  y                = "padj",
  xlab             = bquote(~Log[2]~ "fold change"),
  ylab             = bquote(~-Log[10]~ "adjusted p-value"),
  FCcutoff         = Inf,
  pCutoff          = 0.05,
  cutoffLineWidth  = 0,
  xlim             = c(-4, 4),
  ylim             = c(0, 4),
  colCustom        = keyvals,
  pointSize        = 3.5,
  labSize          = 7,
  boxedLabels      = TRUE,
  drawConnectors   = TRUE,
  widthConnectors  = 0.6,
  colConnectors    = "black",
  title            = NULL,
  subtitle         = NULL,
  caption          = NULL,
  labFace          = "bold"
)

volcano_MC1_MC2

ggsave(
  filename = "../results/figures/poster_volcano_MC1_MC2.png",
  plot     = volcano_MC1_MC2,
  width    = 11, height = 7, units = "in", dpi = 300
)
```

## Control + Exercise vs Control (MC3 vs MC4)

```{r volcano-mc3-mc4}
# no genes pass adjusted p threshold; exercise alone does not drive single-gene changes in bulk striatum
EnhancedVolcano(
  MC3_MC4$shrunk,
  lab      = rownames(MC3_MC4$shrunk),
  x        = "log2FoldChange",
  y        = "padj",
  pCutoff  = 0.05,
  FCcutoff = 1,
  title    = "Control + Exercise vs Control",
  subtitle = NULL
)
```

---

# Pathway Enrichment (FGSEA)

```{r gene-sets}
msig <- msigdbr(
  db_species = "MM",
  species    = "Mus musculus",
  collection = "M5"
) %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(x = .$gene_symbol, f = .$gs_name)
```

```{r fgsea-function}
run_fgsea <- function(res_raw, pathways) {
  res_raw <- res_raw[!is.na(res_raw$stat) & !is.na(res_raw$gene_name), ]
  # deduplicate: keep highest |stat| per gene symbol
  res_raw <- as.data.frame(res_raw) %>%
    dplyr::group_by(gene_name) %>%
    dplyr::slice_max(order_by = abs(stat), n = 1) %>%
    dplyr::ungroup()
  ranks <- setNames(res_raw$stat, res_raw$gene_name)
  fgseaMultilevel(pathways = pathways, stats = ranks)
}
```

```{r run-fgsea}
fg_MC1_MC2 <- run_fgsea(MC1_MC2$raw, msig)
fg_MC3_MC4 <- run_fgsea(MC3_MC4$raw, msig)
```

## Pathway Overlap

```{r pathway-overlap}
sig_MC1_MC2 <- fg_MC1_MC2 %>% filter(padj < 0.05)
sig_MC3_MC4 <- fg_MC3_MC4 %>% filter(padj < 0.05)
shared_pathways <- intersect(sig_MC1_MC2$pathway, sig_MC3_MC4$pathway)

fg_MC1_MC2 %>%
  filter(pathway %in% shared_pathways) %>%
  dplyr::select(pathway, NES, padj) %>%
  arrange(desc(abs(NES)))
```

## FGSEA Dotplot — Selected Transcription Pathways

```{r fgsea-dotplot}
selected_patterns <- c(
  "GOMF_TRANSCRIPTION_REGULATOR_ACTIVITY",
  "GOBP_POSITIVE_REGULATION_OF_TRANSCRIPTION_BY_RNA_POLYMERASE_II",
  "GOCC_TRANSCRIPTION_REGULATOR_COMPLEX",
  "GOMF_CHROMATIN_BINDING"
)

fg_df <- as.data.frame(fg_MC1_MC2)
match_idx   <- unique(unlist(lapply(selected_patterns, function(pat) {
  grep(pat, fg_df$pathway, ignore.case = TRUE, value = FALSE)
})))
# remove any unexpected extra rows (RNA Pol III)
match_idx <- match_idx[-3] 

sel_df <- fg_df[match_idx, ] %>%
  mutate(
    pathway_clean = pathway %>%
      gsub("^GOBP_|^GOCC_|^GOMF_", "", .) %>%
      gsub("_", " ", .) %>%
      stringr::str_to_title()
  )

# manual short labels for plot
sel_df$pathway_clean <- c(
  "Transcription\nRegulator Activity",
  "Positive Regulation of\nTranscription (RNA Pol II)",
  "Transcription \nRegulator Complex",
  "Chromatin \nBinding"
)
sel_df$pathway_clean <- reorder(sel_df$pathway_clean, sel_df$NES)

p_sel <- ggplot(sel_df, aes(x = NES, y = pathway_clean)) +
  geom_point(aes(color = padj),size=15)+
scale_color_gradient(
  low = "#FDB863",
  high = "#B2182B",
  trans = "reverse",
  breaks = c(0.003, 0.0015, 0.00075, 0.0003),
  labels = c("0.003", "0.0015", "0.00075", "0.0003"))+
scale_size_continuous(range = c(12, 20), breaks = pretty(-log10(sel_df$padj))) +
  scale_x_continuous(limits = c(0, max(sel_df$NES) * 1.05)) +
  coord_cartesian(xlim = c(1.35, 1.60))+
  labs(
    x = "Normalized Enrichment Score (NES)",
    y = NULL,
    color = "Adjusted p-value",
    size = NULL
  ) +
  guides(
  color = guide_colorbar(barwidth = 12, barheight = 1)
)+
  theme_classic(base_size = 14) +
  theme(
    aspect.ratio = 0.35,
  axis.text.y = element_text(size = 17, face = "bold", color = "black"),
  axis.title.x = element_text(size = 17, face = "bold", color = "black"),
  axis.text = element_text(size = 17, color = "black"),
  legend.position = "top",
  legend.direction = "horizontal",
  legend.text = element_text(size = 12, color = "black"),
  legend.title = element_text(size = 17, face = "bold", color = "black"),
)
p_sel

ggsave(
  "../results/figures/poster_FGSEA_selected_transcription_paths.png",
  p_sel, width = 10, height = 5, dpi = 300
)
```

---

# TPM Normalization & Candidate Gene Expression

## Compute TPM

```{r compute-tpm}
metadata <- metadata %>% tibble::rownames_to_column(var = "Sample")

count_cols <- setdiff(colnames(counts_df), c("gene_id", "gene_name", "length"))

# fetch gene lengths via biomart
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
bm <- getBM(
  attributes = c("ensembl_gene_id", "start_position", "end_position"),
  filters    = "ensembl_gene_id",
  values     = counts_df$gene_id,
  mart       = ensembl
) %>%
  mutate(length = abs(end_position - start_position) + 1) %>%
  dplyr::select(ensembl_gene_id, length)
colnames(bm)[1] <- "gene_id"

counts_df <- left_join(counts_df, bm, by = "gene_id") %>%
  filter(!is.na(length)) %>%
  mutate(length_kb = length / 1000)

raw_counts_mat <- counts_df %>%
  dplyr::select(all_of(count_cols)) %>%
  as.matrix()
rownames(raw_counts_mat) <- counts_df$gene_id

rpk             <- raw_counts_mat / counts_df$length_kb
scaling_factors <- colSums(rpk, na.rm = TRUE) / 1e6
tpm_mat         <- sweep(rpk, 2, scaling_factors, FUN = "/")

tpm_df <- as.data.frame(tpm_mat) %>%
  tibble::rownames_to_column(var = "gene_id") %>%
  pivot_longer(cols = -gene_id, names_to = "Sample", values_to = "TPM")

gene_id_map <- counts_df %>% dplyr::select(gene_id, gene_name)
tpm_df      <- left_join(tpm_df, gene_id_map, by = "gene_id")
```

## TPM Boxplots

```{r tpm-setup}
genes_of_interest <- c(
  "Fosl2", "Egr1", "Nr4a1", "Nr4a3", "Htr7", "Prlr", "Arhgef16", "Ormdl2",
  "Fos", "Junb", "Bdnf", "Arc", "Camk2a", "Mapk3", "Gfap", "Rbfox3",
  "Dusp1", "Dusp6", "Slc6a3", "Th", "Ddc", "Slc18a2", "Drd1", "Drd2",
  "Ppp1r1b", "Adcy5", "Elk1", "Jun", "Jund", "Atf7",
  "Prkca", "Prkcb", "Prkcg", "Prkcd", "Prkaca",
  "Nr4a2", "Ngf", "Pvr", "Gria1", "Gria2", "Slc2a1", "Thbs1",
  "Maoa", "Maob", "Comt", "Bsg","mt-Co1","mt-Co2","mt-Co3", "Cox4i1", "Cox5a", "Cox6a1","Cox7a1", "Adrb1","Adrb2","Adrb3","Adra1d","Adra2a","Adra2c","Adra1a","Adra1b","Adra2b"
)

group_levels <- c(
  "control_no_exercise", "control_exercise",
  "MPTP_no_exercise",    "MPTP_exercise"
)

group_labels <- c(
  "control_no_exercise" = "Saline",
  "control_exercise"    = "Saline + Exercise",
  "MPTP_no_exercise"    = "MPTP",
  "MPTP_exercise"       = "MPTP + Exercise"
)

plot_df <- tpm_df %>%
  filter(gene_name %in% genes_of_interest) %>%
  left_join(metadata, by = "Sample") %>%
  mutate(
    logTPM      = log2(TPM + 1),
    group       = factor(group, levels = group_levels),
    group_label = factor(group_labels[as.character(group)],
                         levels = unname(group_labels))
  )

comparisons_tpm <- list(
  c("MPTP", "MPTP + Exercise"),
  c("Saline", "Saline + Exercise")
)

dir.create("../results/figures/tpm_boxplots", recursive = TRUE, showWarnings = FALSE)
```

```{r tpm-faceted, fig.width=14, fig.height=14}
# faceted overview — all genes in one plot
tpm_facet <- ggplot(plot_df, aes(x = group_label, y = logTPM, fill = group_label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.15, size = 1.5) +
  facet_wrap(~ gene_name, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 14) +
  labs(x = NULL, y = "log2(TPM + 1)", title = "TPM-normalized expression by group") +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

tpm_facet

ggsave(
  filename = "../results/figures/tpm_boxplots/tpm_all_genes_faceted.png",
  plot     = tpm_facet,
  width    = 14, height = 14, units = "in", dpi = 300
)
```

```{r tpm-individual}
# individual plot per gene saved to tpm_boxplots/
for (g in genes_of_interest) {
  
  df_g <- plot_df %>% filter(gene_name == g)

  p <- ggplot(df_g, aes(x = group_label, y = logTPM, fill = group_label)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.15, size = 2) +
    stat_compare_means(
      comparisons = comparisons_tpm,
      method      = "t.test",
      label       = "p.signif",
      hide.ns     = FALSE,
      tip.length  = 0.02
    ) +
    stat_compare_means(
      method   = "anova",
      label.x  = 3.5,
      label.y  = max(df_g$logTPM, na.rm = TRUE) * 1.5
    ) +
    theme_classic(base_size = 14) +
    labs(
      title = paste0(g, " (TPM-normalized)"),
      x     = NULL,
      y     = "log2(TPM + 1)"
    ) +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )

  ggsave(
    filename = paste0("../results/figures/tpm_boxplots/tpm_", g, ".png"),
    plot     = p, width = 7, height = 6, units = "in", dpi = 300
  )
}
```

---

# Interaction Analysis

## Two-Way ANOVA

```{r two-way-anova}
interaction_genes <- c("Fosl2", "Nr4a1", "Nr4a3", "Ormdl2", "Egr1")

results_list <- lapply(interaction_genes, function(g) {
  df_g  <- plot_df %>% filter(gene_name == g)
  model <- lm(logTPM ~ lesion * exercise, data = df_g)
  an    <- anova(model)

  data.frame(
    Gene                  = g,
    Lesion_p              = an["lesion",           "Pr(>F)"],
    Exercise_p            = an["exercise",          "Pr(>F)"],
    Interaction_p         = an["lesion:exercise",   "Pr(>F)"],
    Exercise_in_MPTP_p    = t.test(logTPM ~ exercise, data = df_g %>% filter(lesion == "MPTP"))$p.value,
    Exercise_in_Control_p = t.test(logTPM ~ exercise, data = df_g %>% filter(lesion == "control"))$p.value
  )
})

results_table <- bind_rows(results_list) %>%
  mutate(across(-Gene, ~ signif(.x, 3)))

colnames(results_table) <- c(
  "Gene", "Lesion (main effect)", "Exercise (main effect)",
  "Lesion × Exercise (interaction)", "Exercise effect in MPTP", "Exercise effect in Control"
)

results_table
```


---

# VST Boxplots

```{r vst-setup}
# fit group-factor model and run LRT for interaction
metadata$group <- factor(metadata$group, levels = group_levels)

dds_group <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = metadata,
  design    = ~ group
)
dds_group <- DESeq(dds_group)

dds_int <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = metadata,
  design    = ~ lesion * exercise
)
dds_int <- DESeq(dds_int, test = "LRT", reduced = ~ lesion + exercise)

res_lrt          <- results(dds_int)
res_ex_in_MPTP   <- results(dds_group, contrast = c("group", "MPTP_exercise",    "MPTP_no_exercise"))
res_ex_in_ctrl   <- results(dds_group, contrast = c("group", "control_exercise", "control_no_exercise"))
shrink_ex_in_MPTP <- lfcShrink(dds_group,
                                contrast = c("group", "MPTP_exercise", "MPTP_no_exercise"),
                                type     = "ashr")

vst_mat <- assay(vst(dds_group, blind = FALSE))

# ensembl -> symbol lookup for vst matrix
sym2ens <- gene_map %>%
  dplyr::filter(!is.na(SYMBOL)) %>%
  dplyr::distinct(SYMBOL, .keep_all = TRUE) %>%
  dplyr::select(SYMBOL, ENSEMBL)

p_to_stars <- function(p) {
  if (is.na(p))    return("ns")
  if (p < 0.001)   return("***")
  if (p < 0.01)    return("**")
  if (p < 0.05)    return("*")
  return("ns")
}

dir.create("../results/figures/vst_boxplots", recursive = TRUE, showWarnings = FALSE)
```

```{r vst-plot-function}
# single-gene VST boxplot with DESeq2-derived pairwise annotations
plot_vst_gene <- function(symbol) {
  # safe group label mapping (used in many chunks)
  group_labels <- c(
    "control_no_exercise" = "Saline",
    "control_exercise"    = "Saline + Exercise",
    "MPTP_no_exercise"    = "MPTP",
    "MPTP_exercise"       = "MPTP + Exercise"
  )

  # map SYMBOL -> ENSEMBL (safe)
  ens <- sym2ens$ENSEMBL[match(symbol, sym2ens$SYMBOL)]
  if (is.na(ens) || length(ens) == 0) {
    warning(sprintf("No ENSEMBL mapping for symbol '%s' — plot will still be produced but brackets will show ns", symbol))
  }
  if (!is.na(ens) && !(ens %in% rownames(vst_mat))) {
    stop(sprintf("Mapped ENSEMBL '%s' for symbol '%s' not present in vst matrix", ens, symbol))
  }

  # build plotting dataframe
  df <- data.frame(
    Sample = colnames(vst_mat),
    vst    = if (!is.na(ens)) as.numeric(vst_mat[ens, , drop=TRUE]) else NA_real_,
    metadata,
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      group_label = factor(group_labels[as.character(group)], levels = unname(group_labels))
    )

  # helper to fetch DESeq2 padj (fallback to pvalue)
  get_p <- function(res, id) {
    if (is.null(res) || is.na(id) || !(id %in% rownames(res))) return(NA_real_)
    p <- res[id, "padj"]
    if (!is.na(p)) return(as.numeric(p))
    pv <- res[id, "pvalue"]
    if (!is.na(pv)) return(as.numeric(pv))
    return(NA_real_)
  }

  # DESeq2 contrast p-values (exercise effect within groups)
  p_ex_MPTP <- get_p(res_ex_in_MPTP, ens)   # exercise effect in MPTP
  p_ex_ctrl <- get_p(res_ex_in_ctrl,  ens)   # exercise effect in control (Saline)

  # shrunken LFC label if available
  lfc_label <- NA
  if (!is.null(shrink_ex_in_MPTP) && !is.na(ens) && ens %in% rownames(shrink_ex_in_MPTP)) {
    lv <- shrink_ex_in_MPTP[ens, "log2FoldChange"]
    if (!is.na(lv)) lfc_label <- signif(lv, 3)
  }

  sub_txt <- paste0(
    "Exercise (MPTP) FDR = ", ifelse(is.na(p_ex_MPTP), "NA", signif(p_ex_MPTP, 2)),
    if (!is.na(lfc_label)) paste0("  |  shrunken LFC = ", lfc_label) else ""
  )

  # dynamic y spacing
  y_max <- max(df$vst, na.rm = TRUE)
  y_min <- min(df$vst, na.rm = TRUE)
  span  <- ifelse(is.finite(y_max - y_min) && (y_max - y_min) > 0, y_max - y_min, 1)
  gap   <- span * 0.12

  # build pval table using the same label strings shown in the plot
  pval_df <- data.frame(
    group1     = c("MPTP", "Saline"),
    group2     = c("MPTP + Exercise", "Saline + Exercise"),
    y.position = c(y_max + gap * 0.6, y_max + gap * 1.2),
    p          = c(p_ex_MPTP, p_ex_ctrl),
    p.signif   = c(p_to_stars(p_ex_MPTP), p_to_stars(p_ex_ctrl)),
    label      = c(
      ifelse(is.na(p_ex_MPTP), "ns", format.pval(p_ex_MPTP, digits = 2)),
      ifelse(is.na(p_ex_ctrl),  "ns", format.pval(p_ex_ctrl,  digits = 2))
    ),
    stringsAsFactors = FALSE
  )

  # plot
  p <- ggplot(df, aes(x = group_label, y = vst, fill = group_label)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.65, width = 0.6, color = "black") +
    geom_jitter(aes(color = group_label), width = 0.12, size = 2, show.legend = FALSE) +
    scale_fill_brewer(palette  = "Set2") +
    scale_color_brewer(palette = "Dark2") +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.15))) +
    coord_cartesian(ylim = c(y_min, y_max + gap * 2), clip = "off") +
    stat_pvalue_manual(
      pval_df,
      label      = "p.signif",   # change to "label" to show numeric adjusted p-value
      xmin       = "group1",
      xmax       = "group2",
      y.position = "y.position",
      tip.length = 0.02,
      size       = 5,
      bracket.size = 0.8
    ) +
    theme_classic(base_size = 14) +
    labs(title = symbol, subtitle = sub_txt, x = NULL, y = "VST expression") +
    theme(
      legend.position = "none",
      plot.title      = element_text(face = "bold", size = 16),
      plot.subtitle   = element_text(size = 10),
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 11),
      axis.title      = element_text(face = "bold"),
      plot.margin     = margin(t = 15, r = 30, b = 10, l = 10)
    )

  return(p)
}
```

```{r vst-faceted, fig.width=14, fig.height=14}
# build tidy df for all genes of interest using vst values
vst_long <- lapply(genes_of_interest, function(g) {
  ens <- sym2ens$ENSEMBL[sym2ens$SYMBOL == g]
  if (length(ens) == 0 || !ens[1] %in% rownames(vst_mat)) return(NULL)
  data.frame(
    Sample    = colnames(vst_mat),
    vst       = as.numeric(vst_mat[ens[1], ]),
    gene_name = g,
    stringsAsFactors = FALSE
  )
}) %>%
  bind_rows() %>%
  left_join(metadata, by = "Sample") %>%
  mutate(
    group_label = factor(group_labels[as.character(group)],
                         levels = unname(group_labels))
  )

vst_facet <- ggplot(vst_long, aes(x = group_label, y = vst, fill = group_label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.15, size = 1.5) +
  facet_wrap(~ gene_name, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 14) +
  labs(x = NULL, y = "VST expression", title = "VST expression by group") +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

vst_facet

ggsave(
  filename = "../results/figures/vst_boxplots/vst_all_genes_faceted.png",
  plot     = vst_facet,
  width    = 14, height = 14, units = "in", dpi = 300
)
```

```{r vst-individual}
# individual plot per gene saved to vst_boxplots/
for (g in genes_of_interest) {
  p <- tryCatch(plot_vst_gene(g), error = function(e) {
    message("skipping ", g, ": ", e$message); NULL
  })
  if (!is.null(p)) {
    ggsave(
      filename = paste0("../results/figures/vst_boxplots/vst_", g, ".png"),
      plot     = p, width = 7, height = 6, units = "in", dpi = 300
    )
  }
}
```

---

# GSVA

```{r gsva, message=FALSE}
expr_mat <- assay(vsd)

# map ensembl rownames to symbols
gene_map_gsva <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys    = rownames(expr_mat),
  keytype = "ENSEMBL",
  columns = "SYMBOL"
) %>%
  dplyr::filter(!is.na(SYMBOL)) %>%
  dplyr::distinct(ENSEMBL, .keep_all = TRUE)

expr_mat_sym <- expr_mat[gene_map_gsva$ENSEMBL, ]
rownames(expr_mat_sym) <- gene_map_gsva$SYMBOL

# average across duplicated gene symbols
sym            <- rownames(expr_mat_sym)
grp            <- as.integer(factor(sym))
summed         <- rowsum(expr_mat_sym, group = grp)
counts_per_sym <- as.vector(table(sym)[levels(factor(sym))])
expr_mat_final <- sweep(summed, 1, counts_per_sym, FUN = "/")
rownames(expr_mat_final) <- levels(factor(sym))

sel_paths <- c(
  "GOMF_TRANSCRIPTION_REGULATOR_ACTIVITY",
  "GOCC_TRANSCRIPTION_REGULATOR_COMPLEX"
)
gs_list_sel <- msig[intersect(sel_paths, names(msig))]

gsva_scores <- gsva(
  expr        = expr_mat_final,
  gset.idx.list = gs_list_sel,
  method      = "ssgsea",
  kcdf        = "Gaussian",
  abs.ranking = TRUE
)
```

```{r gsva-plot, fig.width=8, fig.height=5}
pretty_map <- c(
  "GOCC_TRANSCRIPTION_REGULATOR_COMPLEX"  = "Transcription Regulator\nComplex",
  "GOMF_TRANSCRIPTION_REGULATOR_ACTIVITY" = "Transcription Regulator\nActivity"
)

gsva_plot_df <- as.data.frame(t(gsva_scores)) %>%
  tibble::rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "pathway", values_to = "score") %>%
  left_join(metadata, by = c("sample" = "Sample")) %>%
  mutate(
    condition = factor(
      case_when(
        lesion == "control" & exercise == "no_exercise" ~ "Saline",
        lesion == "control" & exercise == "exercise"    ~ "Saline + Exercise",
        lesion == "MPTP"    & exercise == "no_exercise" ~ "MPTP",
        lesion == "MPTP"    & exercise == "exercise"    ~ "MPTP + Exercise"
      ),
      levels = c("Saline", "Saline + Exercise", "MPTP", "MPTP + Exercise")
    ),
    pathway_label = pretty_map[pathway]
  )

comparisons <- list(
  c("Saline", "Saline + Exercise"),
  c("MPTP",   "MPTP + Exercise")
)

p_gsva <- ggplot(gsva_plot_df, aes(x = condition, y = score)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.9) +
  geom_jitter(width = 0.08, size = 2, color = "black") +
  facet_wrap(~ pathway_label, scales = "free_y") +
  stat_compare_means(
    comparisons = comparisons,
    method      = "t.test",
    label       = "p.signif",
    hide.ns     = TRUE,
    tip.length  = 0.02
  ) +
  theme_classic(base_size = 15) +
  theme(
    strip.text   = element_text(face = "bold"),
    axis.text.x  = element_text(angle = 30, hjust = 1),
    axis.title.x = element_blank()
  ) +
  labs(x = NULL, y = "GSVA pathway score")

p_gsva

ggsave(
  filename = "../results/figures/gsva_transcription_boxplot.png",
  plot     = p_gsva, width = 8, height = 5, dpi = 300
)
```
## Interaction Panel Plot (Poster)

```{r poster-boxplots-fosl2-nr4a1-nr4a3-egr1, fig.width=10, fig.height=7}
genes_to_show <- c("Fosl2", "Nr4a1", "Nr4a3", "Egr1")

label_map_cond <- c(
  "control_no_exercise" = "Saline",
  "control_exercise"    = "Saline + Exercise",
  "MPTP_no_exercise"    = "MPTP",
  "MPTP_exercise"       = "MPTP + Exercise"
)

plot_df2 <- plot_df %>%
  filter(gene_name %in% genes_to_show) %>%
  mutate(
    group_label = factor(
      label_map_cond[as.character(group)],
      levels = c("Saline", "Saline + Exercise", "MPTP", "MPTP + Exercise")
    )
  )

# interaction p-values from anova above 
interaction_pvals <- data.frame(
  Gene          = c("Fosl2", "Nr4a1", "Nr4a3", "Egr1"),
  p_interaction = c(0.00063, 0.0142, 0.00154, 0.03737)
)

plot_gene_panel <- function(gene) {
  df       <- plot_df2 %>% filter(gene_name == gene)
  y_max    <- max(df$logTPM, na.rm = TRUE)
  y_min    <- min(df$logTPM, na.rm = TRUE)
  y_top    <- y_max + 0.5
  p_inter  <- interaction_pvals$p_interaction[interaction_pvals$Gene == gene]
  sub_txt  <- paste0("Interaction FDR = ", formatC(p_inter, format = "e", digits = 2))
  
  # extract the DESeq2 p-value for the gene
  ens <- sym2ens$ENSEMBL[match(gene, sym2ens$SYMBOL)]
  p_mptp <- res_ex_in_MPTP[ens, "padj"]
  p_ctrl <- res_ex_in_ctrl[ens, "padj"]
  
  pval_df <- data.frame(
    group1 = c("MPTP", "Saline"),
    group2 = c("MPTP + Exercise", "Saline + Exercise"),
    y.position = c(y_max + 0.1, y_max + 0.25),
    p = c(p_mptp, p_ctrl),
    p.signif = c(p_to_stars(p_mptp), 
                 p_to_stars(p_ctrl)))
  ggplot(df, aes(x = group_label, y = logTPM, fill = group_label)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, linewidth = 0.6) +
    geom_jitter(width = 0.12, size = 2.2, color = "black") +
    stat_pvalue_manual(
      pval_df,
      label = "p.signif",
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      tip.length = 0.02,
      size = 6,
      bracket.size = 1) +
    scale_y_continuous(expand = expansion(mult = c(0, 0))) +
    coord_cartesian(ylim = c(y_min - 0.05, y_top), clip = "off") +
    scale_fill_brewer(palette = "Set1") +
    theme_classic(base_size = 17) +
    labs(title = gene, subtitle = sub_txt, x = NULL, y = "Expression (log2 TPM + 1)") +
    theme(
      plot.title    = element_text(face = "bold", size = 22),
      plot.subtitle = element_text(size = 9, margin = margin(b = 6)),
      axis.text.x   = element_text(angle = 30, hjust = 1, size = 14, color = "black"), # angle = 40, 
      axis.title.y  = element_text(face = "bold", size = 14),
      axis.text.y   = element_text(size = 12),
      plot.margin   = margin(t = 50, r = 30, b = 12, l = 12),
      legend.position = "none"
    )
}

plots    <- lapply(genes_to_show, plot_gene_panel)
combined <- plot_grid(plotlist = plots, ncol = 2, align = "hv")
combined

ggsave(
  filename = "../results/figures/poster_panel_boxplot.png",
  plot = combined, width = 9, height = 10, units = "in", dpi = 300
)
```


---

# Candidate Gene Selection for qPCR

```{r candidate-genes}
fg_sig <- fg_MC1_MC2 %>% filter(padj < 0.05)

# genes in leading edge of any significant pathway
leading_edge_genes <- as.data.frame(fg_sig) %>%
  dplyr::select(pathway, leadingEdge) %>%
  tidyr::unnest(leadingEdge) %>%
  rename(gene = leadingEdge)

le_genes_unique <- unique(leading_edge_genes$gene)

# significant DEGs
deg_sig <- as.data.frame(MC1_MC2$shrunk) %>%
  filter(padj < 0.05) %>%
  pull(gene_name) %>%
  unique()

# overlap: both DEG and in a leading edge
overlap_genes <- intersect(le_genes_unique, deg_sig)
message("Overlap genes (DEG ∩ leading edge): ", length(overlap_genes))
message("Fraction of DEGs in LE: ", round(length(overlap_genes) / length(deg_sig), 3))

candidate_qpcr <- as.data.frame(MC1_MC2$shrunk) %>%
  filter(gene_name %in% overlap_genes) %>%
  arrange(padj) %>%
  dplyr::select(gene_name, log2FoldChange, padj)

# annotate with pathway membership
qpcr_annotation <- leading_edge_genes %>%
  filter(gene %in% candidate_qpcr$gene_name) %>%
  group_by(gene) %>%
  summarise(
    leading_edge_pathways = paste(unique(pathway), collapse = "; "),
    n_pathways            = n_distinct(pathway),
    .groups               = "drop"
  )

candidate_qpcr_annotated <- candidate_qpcr %>%
  left_join(qpcr_annotation, by = c("gene_name" = "gene")) %>%
  mutate(
    leading_edge_pathways_clean = leading_edge_pathways %>%
      gsub("^GOBP_|^GOCC_|^GOMF_", "", .) %>%
      gsub("_", " ", .)
  )

candidate_qpcr_annotated
```

---

# FGSEA Enrichment Plots

```{r fgsea-enrichment-plots}
# rebuild ranks for plotEnrichment
get_ranks <- function(res_raw) {
  df <- as.data.frame(res_raw) %>%
    filter(!is.na(stat) & !is.na(gene_name)) %>%
    group_by(gene_name) %>%
    slice_max(order_by = abs(stat), n = 1) %>%
    ungroup()
  setNames(df$stat, df$gene_name)
}

ranks <- get_ranks(MC1_MC2$raw)

selected_paths <- c(
  "GOMF_TRANSCRIPTION_REGULATOR_ACTIVITY",
  "GOCC_TRANSCRIPTION_REGULATOR_COMPLEX"
)
selected_paths <- intersect(selected_paths, names(msig))

dir.create("../results/figures/fgsea_enrichment_plots", recursive = TRUE, showWarnings = FALSE)

for (pid in selected_paths) {
  r    <- fg_MC1_MC2[fg_MC1_MC2$pathway == pid, ]
  sub  <- sprintf("NES = %.2f; padj = %.3g; size = %d", r$NES, r$padj, r$size)
  p    <- plotEnrichment(msig[[pid]], stats = ranks) +
    labs(title = pid, subtitle = sub) +
    theme_minimal(base_size = 14)
  ggsave(
    filename = paste0("../results/figures/fgsea_enrichment_plots/", pid, ".png"),
    plot     = p, width = 7, height = 4, dpi = 300
  )
}
```

---

# Session Info

```{r session-info}
sessionInfo()
```